\section{Graphs}

Notation: Graph $G = (V, E)$, where $V$ is set of vertices, $E$ is set of edges, and $|E|$ is number of edges. \textbf{Note:} Sometimes people use $E$ to mean number of edges.

\subsubsection*{Adjacency matrices and lists}

\textbf{Adjacency matrix}: A 2D matrix where $A_{i,j}$ is the number of edges going from $i$ to $j$ (or the number of edges going from $j$ to $i$, depending on preference).\\
Space required is $\Theta(n^2)$.

\textbf{Adjacency list}: An array where the indices represent vertices and the value at index $i$ is a list of all the vertices adjacent to vertex $i$ (maps vertices to their neighbors).\\
Space required is $\Theta(m + n)$, where $m$ is number of edges. $m$ is bounded by $n^2$ (assuming you can only have one edge from one vertex to another)

If the matrix is sparse (not too many edges), adjacency lists are preferred.

Time complexity comparison:\\
\begin{tabular}{|r|c|c|}
     \hline
     & Adjacency matrices & Adjacency lists \\
     \hline
     Check if two vertices adjacent & $\Theta(1)$ & $O(n)$ \\
     \hline
     Find all edges & $\Theta(n^2)$ & $\Theta(m + n)$ \\
     \hline
     Find Eulerian cycle & $\Theta(n^2)$ & $\Theta(m)$ \\
     \hline
\end{tabular}

\subsection{Eulerian Cycles}

\subsubsection*{Terminology}

\textbf{Walk}: A sequence of vertices and edges on a graph that you traverse.

\textbf{Trail}: A walk in which all edges are distinct.

\textbf{Path}: A trail in which all vertices (and edges) are distinct. \hl{TODO figure out the right definition, because by this definition, an Eulerian "path" wouldn't be a real path since vertices repeat}

\textbf{Eulerian path} (a.k.a. Eulerian trail): A trail that visits every edge exactly once (may revisit vertices). The start and end vertices maybe different.

\textbf{Eulerian cycle} (a.k.a. Eulerian circuit, Eulerian tour): An Eulerian path which starts and ends at the same vertex, i.e. a cycle that uses each edge exactly once.

\textbf{Degree}: The degree of a vertex is how many edges are incident to it (equivalently, how many vertices are adjacent to it). A loop contributes 2 to the degree, one for each endpoint.

\textbf{Strongly connected} (like connectedness but for directed graphs): Every vertex is reachable from every other vertex.

\noindent\rule{\textwidth}{1pt}

\subsection*{}

\textbf{Condition for having Eulerian cycle}

An undirected, connected graph $G = (V, E)$ has an \ul{Eulerian cycle if and only if every vertex has even degree}.

Similarly, for directed graphs, the graph must be \emph{strongly} connected and the in degree of each vertex has to equal its out degree (need one edge to go to each vertex and another to go out from that vertex).

\textbf{Condition for having Eulerian path}

An undirected, connected graph $G = (V, E)$ has an Eulerian path if and only if exactly zero or two vertices have odd degree (not possible for only one vertex to have odd degree).

Similarly, for directed graphs, the graph must be \emph{strongly} connected and either every the in degree equals the out degree for every vertex OR there is one vertex whose in degree is 1 greater than its out degree, there is one vertex whose in degree is 1 less than its out degree, and all the other vertices have equal in degrees and out degrees.
