\section{Parallel Merging}

Merging 2 lists of sizes $M$ and $N$ on a sequential computer takes $M + N - 1$ comparisons ($2N - 1$ if same size)

\subsubsection*{Merging with 2 processors}

One processor starts comparing elements at the beginning, the other starts comparing elements at the end, meet in the middle.

Works on machine with Exclusive Read Exclusive Write (EREW).

Takes $\frac{M + N}{2}$ comparisons ($N$ if both lists are same size).

Speedup is $\displaystyle \frac{2N - 1}{N} \approx 2 = P$\\
Efficiency is $\sim 1$

Unfortunately, this algorithm not easily generalizable to more processors.

\subsubsection*{Merging with $N$ processors}

This algorithm seems to require both lists being the same size.

Assign one processor to every element in one of the arrays, then have each processor do binary search to figure out where to place its assigned element.

Requires Concurrent Reads (CREW) because every processor will do binary search at the same time.

Takes $\sim \lg N$ comparisons.

Speedup: $\displaystyle \sim \frac{2P}{\lg P}$, Efficiency: $\displaystyle \sim \frac{2}{\lg P}$

\subsubsection*{Merging when $P < N$}

Watch Rec 22.1 around 13:00

Requires Concurrent Read Exclusive Write (CREW) machine.

Algorithm:
\begin{enumerate}
    \item In both arrays, mark elements $\displaystyle \floor{\frac{2N}{P}}, 2\floor{\frac{2N}{P}}, 3\floor{\frac{2N}{P}}, ...$ (these elements are sort of separators creating $\frac P 2$ groups of elements in each array)
    \item In parallel, binary search each marked element into the other list, creating about $P$ small merging problems. This takes $\sim \lg N$ comparisons.
    \item In parallel, do each small merging problem, assigning two processors to each problem. This takes $\sim \frac{2N}{P}$ comparisons.
\end{enumerate}

Total comparisons: $\displaystyle \sim \frac{2N}{P} + \lg N$

To be efficient, need the number of comparisons to be $\displaystyle O(\frac{2N}{P})$, which requires $N = \Omega(P \lg P)$.

\subsection{Parallel Mergesort}

At level $i$, do $2^i - 1$ comparisons.

Need $\frac N 2$ processors.

Total comparisons: $\sim 2N$

Speedup: $\displaystyle \sim \frac{\lg P}{2}$, Efficiency: $\displaystyle \sim \frac{\lg P}{2P}$

\subsubsection*{Mergesort for Large $N$ (first try)}

Assume $N \geq P$

\begin{enumerate}
    \item Assign $\frac N P$ numbers per processor. Each processor sorts its numbers sequentially ($\displaystyle \frac N P \lg\paren{\frac N P}$ for each processor).
    \item In parallel, pairwise merge lists until there is one sorted list ($\frac{2N}{P}$ for first level from bottom, $\frac{4N}{P}$ for second level from bottom, $\frac{8N}{P}$ for next level, and so on).
\end{enumerate}
Total comparisons for merges (not counting sorting at the bottom): $\displaystyle \sim 2N$\\
Total comparisons (including sorting at bottom): $\displaystyle \sim \frac{N \lg N}{P} + 2N$

To be efficient, we need $\displaystyle N = 2^{\Omega(P)}$

$N$ needs to be exponential in the number of processors, so this is only efficient for very very large $N$. Can improve by using parallel merges:

\subsubsection*{Mergesort using Parallel Merge}

Assume $P = \frac N 2$

The first level from the bottom requires 2 processors and does $\lg 2$ comparisons, second level requires 4 processors and does $\lg 4$ comparisons, third level requires 8 and does $\lg 8$ comparisons, and so on.

Total comparisons: $\displaystyle \sim \frac{(\lg N)^2}{2}$

Speedup: $\displaystyle \sim \frac{4P}{\lg P}$, Efficiency: $\displaystyle \sim \frac{4}{\lg P}$

\subsubsection*{Mergesort for Large $N$ (second try)}

Assume $N \geq P$

Same as before, but this time, merge in parallel (requires CREW)

\begin{enumerate}
    \item Assign $\frac N P$ numbers per processor. Each processor sorts its numbers sequentially ($\displaystyle \frac N P \lg\paren{\frac N P}$ for each processor).
    \item In parallel, pairwise merge lists until there is one sorted list ($\frac{N}{P} + \lg 2$ for first level from bottom, $\frac{N}{P} + \lg 4$ for second level from bottom, $\frac{N}{P} + \lg 8$ for next level, and so on).
\end{enumerate}

Total comparisons for merges (not counting sorting at the bottom): $\displaystyle \sim \frac{N \lg P}{P} + \frac{(\lg P)^2}{2}$\\
Total comparisons (including sorting at bottom): $\displaystyle \sim \frac{N \lg N}{P} + \frac{(\lg P)^2}{2}$

To be efficient, need $N = \Omega(P \lg P)$ (doesn't need to be huge, unlike before)

\subsubsection*{Sorting with Optimal Merging Algorithm}

Can merge two lists each of size $N$ on CREW PRAM with $P \leq N$ processors in $\sim 2\frac N P + \lg \lg P$ comparison steps.

So can sort on CREW PRAM with $\displaystyle \sim \frac{N \lg N}{P} + (\lg P)\lg \lg P$ comparison steps.

This algorithm is efficient for $N = \Omega(P \lg \lg P)$

\subsection*{Pipelining}

One processor can start merging as soon as processors on bottom levels are done with elements at the start, even if they're not fully finished.

With pipelining, mergesort is $\displaystyle \Theta\paren{\frac{N \lg N}{P}}$ if $N \geq P$

Efficient for $N = \Omega(P)$
